Bayesian multivariate skew-normal finite mixture model for analysis of infant development trajectories

### 2019-06-25 manuscript/manuscript.tex
section 3

%  ALWAYS USE THE referee OPTION WITH PAPERS SUBMITTED TO BIOMETRICS!!!
%  You can see what your paper would look like typeset by removing
%  the referee option.  Because the typeset version will be in two
%  columns, however, some of your equations may be too long. DO NOT
%  use the \longequation option discussed in the user guide!!!  This option
%  is reserved ONLY for equations that are impossible to split across 
%  multiple lines; e.g., a very wide matrix.  Instead, type your equations 
%  so that they stay in one column and are split across several lines, 
%  as are almost all equations in the journal.  Use a recent version of the
%  journal as a guide. 
%  
\documentclass[useAMS,referee]{biom}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{blindtext}
\usepackage{booktabs}
\usepackage{breqn}
\usepackage{algorithm}
\usepackage{algcompatible}
\usepackage{algpseudocode}
\usepackage{float}
\usepackage{dsfont}

@@

A primary goal of the Nurture study is to identify clusters of infants characterized by distinct motor development trajectories. To address this aim, we propose a flexible finite mixture model that accommodates relevant features of the data, such as skewness and dependence among the responses. To this end, let $\mathbf{y}_{i}=(y_{i1},\ldots,y_{iJ})^T$ be a $J \times 1$ vector of responses (i.e., Baley scores) for subject $i$ $(i=1,\ldots,n)$. For the analysis of the Nurture data, we propose a finite mixture model of the form
\begin{equation}
f(\mathbf{y}_i) = \sum_{k = 1}^{K} \pi_{ik} f(\mathbf{y}_i|\boldsymbol\theta_k),
\end{equation}
where $\boldsymbol\theta_k$ is the set of parameters specific to cluster $k$ ($k = 1,...,K$) and $\pi_{ik}$ is a subject-specific mixing weight representing the probability that subject $i$ belongs to cluster $k$. For now we assume that $K$ is fixed; in Section 4, we discuss model selection strategies for choosing the optimal value of $K$.

To facilitate posterior inference, we introduce a latent cluster indicator variable $z_i$ taking the value $k \in \{1,...,K\}$ with probability $\pi_{ik}$. Conditional on $z_i = k$, we assume $\mathbf{y}_{i}$ is distributed as
\begin{equation}
\mathbf{y}_{i}|(z_i=k) \sim MSN_J(\boldsymbol\zeta_{ki},\boldsymbol\alpha_k,\boldsymbol\Omega_k), \label{eq:msndens}
\end{equation}
where $MSN_J(\cdot)$ denotes the $J$-dimensional multivariate skew normal density, $\boldsymbol\zeta_{ki}$ is a $J \times 1$ vector of subject- and cluster-specific location parameters, $\boldsymbol\alpha_k$ is a $J \times 1$ vector of cluster-specific skewness parameters, and $\boldsymbol\Omega_k$ is a $J \times J$ cluster-specific scale matrix that captures dependence among the $J$ responses. The vector $\boldsymbol\alpha_k$ has components $\alpha_{kj}$, $j = 1,...,J$, that control the skewness of outcome $j$ in cluster $k$. When $\boldsymbol\alpha_k = \mathbf{0}$, the MSN distribution reduces to the multivariate normal distribution $N_J(\boldsymbol\zeta_k,\boldsymbol\Omega_k)$, where $\boldsymbol\Omega_k$ is a $J \times J$ covariance matrix.

We can extend model (\ref{eq:msndens}) to the regression setting by modeling $\boldsymbol\zeta_{ki}$ as a function of covariates. Here we adopt a convenient stochastic representation of the MSN density (Azzalini and Dalla Valle, 1996):
\begin{equation}
\mathbf{y}_{i}|(z_i=k,t_i) = \mathbf{X}_i \boldsymbol\beta_k + t_i \boldsymbol\psi_k + \boldsymbol\epsilon_{ki}, \label{eq:msnreg}
\end{equation}
where 
$\mathbf{X}_i$ is a $J \times Jp$ design matrix that includes potential time-varying covariates (e.g., indicators denoting quarterly visits); $\boldsymbol{\beta}_k=(\beta_{k11},\ldots,\beta_{k1p},\ldots,\beta_{kJ1},\ldots,\beta_{kJp})^T$ is a $Jp\times 1$ vector of cluster- and outcome-specific regression coefficients; $t_i\sim N_{[0,\infty)}(0,1)$ is a subject-specific standard normal random variable truncated below by zero; $\boldsymbol\psi_k=(\psi_{k1},\ldots,\psi_{kJ})^T$ is a $J \times 1$ vector of cluster-specific skewness parameters; and $\boldsymbol{\epsilon}_{ki} \sim N_J(\boldsymbol0,\boldsymbol\Sigma_k)$ is a $J\times 1$ vector of error terms. Thus, conditional on $t_i$ and $z_i=k$, $\boldsymbol{y}_i$ is distributed as $N_J(\mathbf{X}_i \boldsymbol\beta_k + t_i \boldsymbol\psi_k, \boldsymbol{\Sigma}_k)$. Marginally (after integrating over $t_i$), $\mathbf{y}_i$ is distributed $MSN_J(\boldsymbol\zeta_{ki}, \boldsymbol\alpha_k, \boldsymbol\Omega_k)$, where through back-transformation 
\begin{eqnarray*}
\boldsymbol\zeta_{ki} &=& \mathbf{X}_i\boldsymbol\beta_k,\\
\boldsymbol\alpha_k &=& \frac{1}{\sqrt{1 - \boldsymbol\psi_k^T 
\boldsymbol\Omega^{-1}_k\boldsymbol\psi_k}} \boldsymbol\omega_k \boldsymbol\Omega^{-1}_k\boldsymbol\psi_k,~~\text{and}\\
\boldsymbol\Omega_k &=& \boldsymbol\Sigma_k  + \boldsymbol\psi_k \boldsymbol\psi_k^T,
\end{eqnarray*}
where $\boldsymbol\omega_k = \text{Diag}(\sqrt{\omega_{k,11}},...,\sqrt{\omega_{k,JJ}})$ is the $J \times J$ diagonal matrix containing the square root of the diagonal entries of $\boldsymbol\Omega_k$. Additional details can be found in Fr\"{u}wirth-Schnatter and Pyne (2010).

Of note, the MSN density can be expressed more compactly in terms of the matrix skew normal (MatSN) density (Chen and Gupta 2005). As we will see in Section 3.6, the matrix representation of the MSN distribution admits convenient conjugate prior distributions for the regression parameters and scale matrices, which in turn leads to efficient Gibbs sampling for posterior inference. Let $\mathbf{Y}_k$ be an ${n_k \times J}$ response matrix with rows $\mathbf{y}_i^T$, $(i = 1,...,n_k)$, where $n_k = \sum_{i = 1}^n {1}_{(z_i = k)}$ is the number of observations in cluster $k$. From equation (\ref{eq:msnreg}), it follows that $\mathbf{Y}_k$ is distributed as
\begin{eqnarray*}
\mathbf{Y}_{k} &\sim& MatSN_{n_k \times J}(\mathbf{M}_k,\boldsymbol\alpha_k,\mathbf{I}_{n_k},\boldsymbol\Omega_k)\\
%\text{vec}(\mathbf{Y}_k) =  (\mathbf{y}^T_1,...,\mathbf{y}^T_{n_k})^T\\
\text{vec}(\mathbf{M}_k) &=& (\boldsymbol\zeta_{k1}^T,...,\boldsymbol\zeta_{kn_k}^T)^T,
\end{eqnarray*}
where $\boldsymbol\zeta_{ki} = \mathbf{X}_i \boldsymbol\beta_k$ as in equation (\ref{eq:msnreg}), $\boldsymbol\alpha_k = (\alpha_{k1},...,\alpha_{kJ})^T$, $\mathbf{I}_{n_k}$ is the $n_k\times n_k$ identity matrix, and $\boldsymbol\Omega_k$ is the $J\times J$ scale matrix defined above in equation (\ref{eq:msndens}). From equation (\ref{eq:msnreg}), it follows that $\mathbf{Y}_{k}$, conditional on the $n_k\times 1$ vector of random effects $\mathbf{t}_{k}$, is jointly distributed in matrix form as
\[
\mathbf{Y}_k | \mathbf{t}_k \sim MatNorm_{n_k \times J}(\mathbf{M}_k, \mathbf{I}_{n_k}, \boldsymbol\Sigma_k),
\]
where $MatNorm_{n_k \times J}(\cdot)$ denotes a $n_k\times J$ matrix normal density, $\text{vec}(\mathbf{M}_k) = \mathbf{X}_{k} \boldsymbol\beta_k + \mathbf{t}_k \otimes \boldsymbol\psi_k$ is an $n_kJ\times 1$ mean vector, $\mathbf{X}_k$ is an $n_kJ \times Jp$ design matrix, $\boldsymbol\beta_k$ is the $(Jp)\times 1$ vector of regression coefficients defined in equation (\ref{eq:msnreg}), and $\boldsymbol\Sigma_k$ is the $J\times J$ conditional covariance of $\boldsymbol\epsilon_{ik}$ given in equation (\ref{eq:msnreg}).

\subsection{Multinomial Regression for the Cluster Indicators}
To accommodate heterogeneity in the cluster-membership probabilities, we model $\pi_{ik}$ as a function of coovariates using a multinomial logit model 
\begin{eqnarray}
\pi_{ik} = \Pr(z_i = k|\mathbf{w}_i) = \frac{e^{\mathbf{w}_i^T \boldsymbol\delta_k}}{\sum_{h = 1}^K e^{\mathbf{w}_i^T \boldsymbol\delta_{h}}},~ k=1,\ldots,K,
\end{eqnarray}
where $\mathbf{w}_i$ is an $r\times 1$ vector of subject-level covariates, $\boldsymbol\delta_k$ is a $r\times 1$ vector of regression parameters associated with membership in cluster $k$. For identifiability purposes, we fix the reference category $k = K$ and set $\boldsymbol\delta_K = \mathbf{0}$. Under this model, $z_i|\mathbf{w}_i \sim Multinom(1,\boldsymbol\pi_i)$, where $\boldsymbol\pi_i = (\pi_{i1},...,\pi_{1K})$. During MCMC estimation, the cluster labels $z_i$ are updated from their multinomial full conditional distribution and used in the remaining MCMC steps as cluster assignments.

\subsection{Conditional MSN Imputation}
To accommodate missing at random (MAR) responses, we propose a convenient imputation algorithm that can be implemented ``online'' as part of the Gibbs sampler. In Section 6, we discuss extensions to allow for non-ignorable missingness. Suppose $\mathbf{y}_i$ has $q_i\in (1,\ldots,J)$ observed values, denoted $\mathbf{y}^{obs}_i$, and $J-q_i$ intermittent missing values, denoted $\mathbf{y}^{miss}_i$. We can make use of the stochastic representation given in equation (\ref{eq:msnreg}) to impute $\mathbf{y}^{miss}_i$ from its conditional multivariate normal distribution given $(z_i,t_i,Y^{obs}_i)$:
\begin{eqnarray}
\mathbf{y}^{miss}_i|(z_i=k,t_i,\mathbf{y}^{obs}_i)&\sim& N_{J-q_i}(\boldsymbol\mu^{cond}_{ki},\boldsymbol\Sigma^{cond}_k),~\text{where}\nonumber\\
\boldsymbol\mu^{cond}_{ki}&=& \boldsymbol\mu^{miss}_{ki} + \boldsymbol\Sigma_{12} \boldsymbol\Sigma_{22}^{-1}(\mathbf{y}^{obs}_i - \boldsymbol\mu^{obs}_{ki}) \label{eq:impute}\\
\boldsymbol\Sigma^{cond}_k&=& \boldsymbol\Sigma_{11} - \boldsymbol\Sigma_{12}\boldsymbol\Sigma_{22}^{-1}\boldsymbol\Sigma_{21},~\text{where}\nonumber
\end{eqnarray}
$\boldsymbol\Sigma_k$ is partitioned into four sub-matrices $\boldsymbol\Sigma_{11},\boldsymbol\Sigma_{12},\boldsymbol\Sigma_{21}, \ \text{and} \ \boldsymbol\Sigma_{22}$ such that $\boldsymbol\Sigma_{11}$ is a $J-q_i \times J-q_i$ matrix containing the rows and columns of $\boldsymbol\Sigma_k$ corresponding to inicies of $\mathbf{y}_{i}$ where missingness occurs. Similarly, $\boldsymbol\Sigma_{12}$ is a $J-q_i \times q_i$ matrix containing the rows of $\boldsymbol\Sigma_k$ that correspond to missing indices of $\mathbf{y}_i$, but columns of $\boldsymbol\Sigma_k$ that correspond to observed indices of $\mathbf{y}_i$. The remaining partitions $\boldsymbol\Sigma_{21}, \ \text{and} \ \boldsymbol\Sigma_{22}$ are defined in the same manner. These results follow from conventional multivariate normal theory. An attractive feature of this imputation algorithm is that it avoids multiplicative run-time scaling in $m$, the number of imputations (Gelman \textit{et al.} 2013; Zhou and Reiter, 2010). Our approach also provides more opportunities to explore the missing data parameter space than does multiple imputation, since missing values are drawn at each MCMC iteration, and often in practice $n_{sim} >> m$, where $n_{sim}$ is the number of MCMC iterations. In Section 4, we conduct simulation studies to demonstrate that imputing the missing MSN responses improves inferences over complete case analysis.

@@

\subsubsection{Prior Specification}
We adopt a fully Bayesian inferential approach and assign prior distributions to all model parameters. Conveniently, all parameters admit conditionally conjugate priors, which greatly improves posterior computation via a data-augmented Gibbs sampler. For the MSN model component, we adopt a conditionally independent prior structure for $\boldsymbol\beta_k$ and $\boldsymbol\Sigma_k$, where $p(\boldsymbol\beta_k,\boldsymbol\Sigma_k) = p(\boldsymbol\Sigma_k)p(\boldsymbol\beta_k|\boldsymbol\Sigma_k)$. We choose the normal-inverse-Wishart distribution for $p(\boldsymbol\beta_k,\boldsymbol\Sigma_k)$ by specifying $\boldsymbol\Sigma_k \sim \text{IW}(\mathbf{V}_k,\nu_k)$ and $\boldsymbol\beta_k|\boldsymbol\Sigma_k \sim \text{N}_{Jp}(\mathbf{b}_k,\mathbf{I}_p \otimes \boldsymbol\Sigma_k)$. We assign the skewness parameters $\boldsymbol\psi_k$ a conjugate $\text{N}_J(\mathbf{m}_k,\mathbf{P}_k)$ prior. However, the updates of $\boldsymbol\beta_k$ and $\boldsymbol\psi_k$ can be combined into one step by defining the $(Jp + J) \times 1$ vector $\boldsymbol\beta^*_k = (\boldsymbol\beta_k,\boldsymbol\psi_k)^T$ for which we assume a $\text{N}_{Jp+J}(\mathbf{b}^*_k,\mathbf{I}_{(p+1)} \otimes \boldsymbol\Sigma_k)$ prior. For the multinomial logit model component, the regression parameters $\boldsymbol\delta_k$ are given a conjugate $\text{N}_r(\mathbf{d}_k, \mathbf{S}_k)$ prior for $k = 1,...,K-1$.

We allow the normal-inverse-Wishart and multinomial hyperparameters to vary across clusters, though they may be shared across clusters in practice. An advantage of allowing for cluster-specific prior parameters is that \textit{a priori} knowledge of development trends can be incorporated into certain clusters while still allowing the parameters of other clusters to be almost entirely determined by the data. Additionally, prior information regarding the effect of certain covariates on development cluster membership can be incorporated in to the model by choosing informative values for $\mathbf{d}_k \ \text{and} \ \mathbf{S}_k$

\subsubsection{Posterior Inference} 
The above prior specification induces closed-form full conditionals that can be efficiently updated as part of a Gibbs sampler outlined below. A programatic sketch of our MCMC algorithm is given in table (\ref{alg:MCMC}). Additional details, including derivations can be found in the Web Appendix.

@@

\paragraph{\textbf{Step 2:} Update of MSN Regression Parameters}

\paragraph{\textbf{Step 3:}P\'olya--Gamma Data Augmentation for $z_i$} The sampler concludes by updating the latent cluster indicators $z_i$ $(i=1,\ldots,n)$ from its multinomial logit full conditional. To facilitate sampling, we adopt an efficient data-augmentation approach introduced by Polson \textit{et al.} (2013), which expresses the inverse-logit function as a mixture P\'olya--Gamma densities. By using P\'olya--Gamma data augmentation for the multinomial model, we obtain a P\'olya--Gamma mixture of experts model, a computationally efficient way to model edge weights (\textbf{cite}). A random variable $w$ is said to follow a P\'olya--Gamma distribution with parameters $b > 0$ and $c \in {\rm I\!R}$ if
\begin{equation}
w \sim \text{PG}(b,c) \stackrel{d}{=} \frac{1}{2\pi^2}\sum_{s=1}^{\infty}\frac{g_s}{(s-1/2)^2 + c^2/(4\pi^2)}, \label{eq:pg1}
\end{equation}
where $g_s \stackrel{iid}{\sim} \text{Ga}(b,1)$ for $s = 1,...,\infty$ . Polson \textit{et al.} (2013) establish that for the $\text{PG}(b,c)$ density, and for $a,\eta \in {\rm I\!R}$,

\begin{equation}
\frac{(e^{\eta})^a}{(1 + e^{\eta})^b} = 2^{-b}e^{\kappa \eta} \int_{0}^{\infty} e^{-w\eta^2/2}p(w|b,c = 0)dw. \label{eq:pg2}
\end{equation}
where $\kappa = a - b/2$. Polson \textit{et al.} also show that the conditional distribution $p(w|b,c)$ results from an ``exponential tilting'' of the $\text{PG}(b,0)$ density, thus

>>>>

%  ALWAYS USE THE referee OPTION WITH PAPERS SUBMITTED TO BIOMETRICS!!!
%  You can see what your paper would look like typeset by removing
%  the referee option.  Because the typeset version will be in two
%  columns, however, some of your equations may be too long. DO NOT
%  use the \longequation option discussed in the user guide!!!  This option
%  is reserved ONLY for equations that are impossible to split across 
%  multiple lines; e.g., a very wide matrix.  Instead, type your equations 
%  so that they stay in one column and are split across several lines, 
%  as are almost all equations in the journal.  Use a recent version of the
%  journal as a guide. 
%  
\documentclass[useAMS,referee]{biom}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{blindtext}
\usepackage{booktabs}
\usepackage{breqn}
\usepackage{algorithm}
\usepackage{algcompatible}
\usepackage{algpseudocode}
\usepackage{float}
\usepackage{dsfont}
\usepackage{amsfonts}

@@

A primary goal of the Nurture study is to identify clusters of infants characterized by distinct motor development trajectories. To address this aim, we propose a flexible finite mixture model that accommodates relevant features of the data, such as skewness and dependence among the responses. For $i = 1,...,n$, let $\mathbf{y}_{i}=(y_{i1},\ldots,y_{iJ})^T$ be a $J \times 1$ vector of responses (i.e., Baley scores) for subject $i$ across the $J$ measurement occasions. For the analysis of the Nurture data, we propose a finite mixture model of the form
\begin{equation}
f(\mathbf{y}_i) = \sum_{k = 1}^{K} \pi_{ik} f(\mathbf{y}_i|\boldsymbol\theta_k),
\end{equation}
where $\boldsymbol\theta_k$ is the set of parameters specific to cluster $k$ ($k = 1,...,K$) and $\pi_{ik}$ is a subject-specific mixing weight representing the probability that subject $i$ belongs to cluster $k$. For now we assume that $K$ is fixed; in Section 4, we discuss model selection strategies for choosing the optimal value of $K$.

To facilitate posterior inference, we introduce a latent cluster indicator variable $z_i$ taking the value $k \in \{1,...,K\}$ with probability $\pi_{ik}$. Conditional on $z_i = k$, we assume $\mathbf{y}_{i}$ is distributed as
\begin{equation}
\mathbf{y}_{i}|(z_i=k) \sim \text{MSN}_J(\boldsymbol\zeta_{ki},\boldsymbol\alpha_k,\boldsymbol\Omega_k), \label{eq:msndens}
\end{equation}
where $\text{MSN}_J(\cdot)$ denotes the $J$-dimensional multivariate skew normal density, $\boldsymbol\zeta_{ki}$ is a $J \times 1$ vector of subject- and cluster-specific location parameters, $\boldsymbol\alpha_k$ is a $J \times 1$ vector of cluster-specific skewness parameters, and $\boldsymbol\Omega_k$ is a $J \times J$ cluster-specific scale matrix that captures dependence among the $J$ responses for subject $i$. The vector $\boldsymbol\alpha_k$ has components $\alpha_{kj}$, $j = 1,...,J$, that control the skewness of outcome $j$ in cluster $k$. When $\boldsymbol\alpha_k = \mathbf{0}$, the MSN distribution reduces to the multivariate normal distribution $\text{N}_J(\boldsymbol\zeta_k,\boldsymbol\Omega_k)$, where $\boldsymbol\zeta_k$ is a $J \times 1$ mean vector and $\boldsymbol\Omega_k$ is a $J \times J$ covariance matrix.

We can extend model (\ref{eq:msndens}) to the regression setting by modeling $\boldsymbol\zeta_{ki}$ as a function of covariates. Here we adopt a convenient stochastic representation of the MSN density (Azzalini and Dalla Valle, 1996):
\begin{equation}
\mathbf{y}_{i}|(z_i=k,t_i) = \mathbf{X}_i \boldsymbol\beta_k + t_i \boldsymbol\psi_k + \boldsymbol\epsilon_{ki}, \label{eq:msnreg}
\end{equation}
where 
$\mathbf{X}_i$ is a $J \times Jp$ design matrix that includes potential time-varying covariates (e.g., indicators denoting quarterly visits); $\boldsymbol{\beta}_k=(\beta_{k11},\ldots,\beta_{k1p},\ldots,\beta_{kJ1},\ldots,\beta_{kJp})^T$ is a $Jp\times 1$ vector of cluster- and outcome-specific regression coefficients; $t_i\sim \text{N}_{[0,\infty)}(0,1)$ is a subject-specific standard normal random variable truncated below by zero; $\boldsymbol\psi_k=(\psi_{k1},\ldots,\psi_{kJ})^T$ is a $J \times 1$ vector of cluster-specific skewness parameters; and $\boldsymbol{\epsilon}_{ki} \sim \text{N}_J(\boldsymbol0,\boldsymbol\Sigma_k)$ is a $J\times 1$ vector of error terms. Thus, conditional on $t_i$ and $z_i=k$, $\boldsymbol{y}_i$ is distributed as $\text{N}_J(\mathbf{X}_i \boldsymbol\beta_k + t_i \boldsymbol\psi_k, \boldsymbol{\Sigma}_k)$. Marginally (after integrating over $t_i$), $\mathbf{y}_i$ is distributed $\text{MSN}_J(\boldsymbol\zeta_{ki}, \boldsymbol\alpha_k, \boldsymbol\Omega_k)$, where through back-transformation 
\begin{eqnarray*}
\boldsymbol\zeta_{ki} &=& \mathbf{X}_i\boldsymbol\beta_k,\\
\boldsymbol\alpha_k &=& \frac{1}{\sqrt{1 - \boldsymbol\psi_k^T 
\boldsymbol\Omega^{-1}_k\boldsymbol\psi_k}} \boldsymbol\omega_k \boldsymbol\Omega^{-1}_k\boldsymbol\psi_k,~~\text{and}\\
\boldsymbol\Omega_k &=& \boldsymbol\Sigma_k  + \boldsymbol\psi_k \boldsymbol\psi_k^T,
\end{eqnarray*}
where $\boldsymbol\omega_k = \text{Diag}(\boldsymbol\Omega_k)^{1/2}$ is the $J \times J$ diagonal matrix containing the square root of the diagonal entries of $\boldsymbol\Omega_k$. Additional details can be found in Fr\"{u}wirth-Schnatter and Pyne (2010).

Of note, the MSN density can be expressed more compactly in terms of the matrix skew normal (MatSN) density (Chen and Gupta 2005). As we will see in Section 3.6, the matrix representation of the MSN distribution admits convenient conjugate prior distributions for the regression parameters and scale matrices, which in turn leads to efficient Gibbs sampling for posterior inference. Let $\mathbf{Y}_k$ be an ${n_k \times J}$ response matrix with rows $\mathbf{y}_i^T$, $(i = 1,...,n_k)$, where $n_k = \sum_{i = 1}^n {1}_{(z_i = k)}$ is the number of observations in cluster $k$. From equation (\ref{eq:msnreg}), it follows that $\mathbf{Y}_k$ is distributed as
\begin{eqnarray*}
\mathbf{Y}_{k} &\sim& \text{MatSN}_{n_k \times J}(\mathbf{M}_k,\boldsymbol\alpha_k,\mathbf{I}_{n_k},\boldsymbol\Omega_k)\\
%\text{vec}(\mathbf{Y}_k) =  (\mathbf{y}^T_1,...,\mathbf{y}^T_{n_k})^T\\
\text{vec}(\mathbf{M}_k) &=& (\boldsymbol\zeta_{k1}^T,...,\boldsymbol\zeta_{kn_k}^T)^T,
\end{eqnarray*}
where $\boldsymbol\zeta_{ki} = \mathbf{X}_i \boldsymbol\beta_k$ as in equation (\ref{eq:msnreg}), $\boldsymbol\alpha_k = (\alpha_{k1},...,\alpha_{kJ})^T$, $\mathbf{I}_{n_k}$ is the $n_k\times n_k$ identity matrix, and $\boldsymbol\Omega_k$ is the $J\times J$ scale matrix defined above in equation (\ref{eq:msndens}). From equation (\ref{eq:msnreg}), it follows that $\mathbf{Y}_{k}$, conditional on the $n_k\times 1$ vector of random effects $\mathbf{t}_{k} = (t_1,...,t_{n_k})^T$, is jointly distributed in matrix form as
\[
\mathbf{Y}_k | \mathbf{t}_k \sim \text{MatNorm}_{n_k \times J}(\mathbf{M}_k, \mathbf{I}_{n_k}, \boldsymbol\Sigma_k),
\]
where $\text{MatNorm}_{n_k \times J}(\cdot)$ denotes a $n_k\times J$ matrix normal density, $\text{vec}(\mathbf{M}_k) = \mathbf{X}_{k} \boldsymbol\beta_k + \mathbf{t}_k \otimes \boldsymbol\psi_k$ is an $n_kJ\times 1$ mean vector, $\mathbf{X}_k$ is an $n_kJ \times Jp$ design matrix, $\boldsymbol\beta_k$ is the $(Jp)\times 1$ vector of regression coefficients defined in equation (\ref{eq:msnreg}), and $\boldsymbol\Sigma_k$ is the $J\times J$ conditional covariance of $\boldsymbol\epsilon_{ik}$ given in equation (\ref{eq:msnreg}).

\subsection{Multinomial Regression for the Cluster Indicators}
To accommodate heterogeneity in the cluster-membership probabilities, we model $\pi_{ik}$ as a function of covariates using a multinomial logit model 
\begin{eqnarray}
\pi_{ik} = \Pr(z_i = k|\mathbf{w}_i) = \frac{\text{e}^{\mathbf{w}_i^T \boldsymbol\delta_k}}{\sum_{h = 1}^K \text{e}^{\mathbf{w}_i^T \boldsymbol\delta_{h}}},~ k=1,\ldots,K,
\end{eqnarray}
where $\mathbf{w}_i$ is an $r\times 1$ vector of subject-level covariates, $\boldsymbol\delta_k$ is a $r\times 1$ vector of regression parameters associated with membership in cluster $k$. For identifiability purposes, we fix the reference category $k = K$ and set $\boldsymbol\delta_K = \mathbf{0}$. Under this model, $z_i|\mathbf{w}_i \sim \text{Multinom}(1,\boldsymbol\pi_i)$, where $\boldsymbol\pi_i = (\pi_{i1},...,\pi_{1K})$. During MCMC estimation, the cluster labels $z_i$ are updated from their multinomial full conditional distribution and used in the remaining MCMC steps as cluster assignments.

\subsection{Conditional MSN Imputation}
To accommodate missing at random (MAR) responses, we propose a convenient imputation algorithm that can be implemented ``online'' as part of the Gibbs sampler. In Section 6, we discuss extensions to allow for non-ignorable missingness. Suppose $\mathbf{y}_i$ has $q_i\in (1,\ldots,J)$ observed values, denoted $\mathbf{y}^{obs}_i$, and $J-q_i$ intermittent missing values, denoted $\mathbf{y}^{miss}_i$. We can make use of the stochastic representation given in equation (\ref{eq:msnreg}) to impute $\mathbf{y}^{miss}_i$ from its conditional multivariate normal distribution given $(z_i,t_i,\mathbf{y}^{obs}_i)$:
\begin{eqnarray}
\mathbf{y}^{miss}_i|(z_i=k,t_i,\mathbf{y}^{obs}_i)&\sim& \text{N}_{J-q_i}(\boldsymbol\mu^{cond}_{ki},\boldsymbol\Sigma^{cond}_k),~\text{where}\nonumber\\
\boldsymbol\mu^{cond}_{ki}&=& \boldsymbol\mu^{miss}_{ki} + \boldsymbol\Sigma_{12} \boldsymbol\Sigma_{22}^{-1}(\mathbf{y}^{obs}_i - \boldsymbol\mu^{obs}_{ki}) \label{eq:impute}\\
\boldsymbol\Sigma^{cond}_k&=& \boldsymbol\Sigma_{11} - \boldsymbol\Sigma_{12}\boldsymbol\Sigma_{22}^{-1}\boldsymbol\Sigma_{21},~\text{and}\nonumber \\
\boldsymbol\Sigma_k &=& \begin{pmatrix} 
	\boldsymbol\Sigma_{k11} & \boldsymbol\Sigma_{k12}\\
	\boldsymbol\Sigma_{k21} & \boldsymbol\Sigma_{k22}
\end{pmatrix}, \ \text{where} \nonumber
\end{eqnarray}
$\boldsymbol\Sigma_{k11}$ is a $J-q_i \times J-q_i$ matrix containing the rows and columns of $\boldsymbol\Sigma_k$ corresponding to inicies of $\mathbf{y}_{i}$ where missingness occurs. Similarly, $\boldsymbol\Sigma_{k12}$ is a $J-q_i \times q_i$ matrix containing the rows of $\boldsymbol\Sigma_k$ that correspond to missing indices of $\mathbf{y}_i$, but columns of $\boldsymbol\Sigma_k$ that correspond to observed indices of $\mathbf{y}_i$. The remaining partitions $\boldsymbol\Sigma_{k21}, \ \text{and} \ \boldsymbol\Sigma_{k22}$ are defined in the same manner. These results follow from conventional multivariate normal theory. An attractive feature of this imputation algorithm is that it avoids multiplicative run-time scaling in $m$, the number of imputations (Gelman \textit{et al.} 2013; Zhou and Reiter, 2010). Our approach also provides more opportunities to explore the missing data parameter space than does multiple imputation, since missing values are drawn at each MCMC iteration, and often in practice $n_{sim} >> m$, where $n_{sim}$ is the number of MCMC iterations (\textbf{find a reference}). In Section 4, we conduct simulation studies to demonstrate that imputing the missing MSN responses improves inferences over complete case analysis.

@@

\subsubsection{Prior Specification}
We adopt a fully Bayesian inferential approach and assign prior distributions to all model parameters. Conveniently, all parameters admit conditionally conjugate priors, which greatly improves posterior computation via a data-augmented Gibbs sampler. For the MSN model component, we adopt a conditionally independent prior structure for $\boldsymbol\beta_k$ and $\boldsymbol\Sigma_k$, where $p(\boldsymbol\beta_k,\boldsymbol\Sigma_k) = p(\boldsymbol\Sigma_k)p(\boldsymbol\beta_k|\boldsymbol\Sigma_k)$. We choose the normal-inverse-Wishart distribution for $p(\boldsymbol\beta_k,\boldsymbol\Sigma_k)$ by specifying $\boldsymbol\Sigma_k \sim \text{IW}(\mathbf{V}_{0k},\nu_{0k})$ and $\boldsymbol\beta_k|\boldsymbol\Sigma_k \sim \text{N}_{Jp}(\mathbf{b}_k,\mathbf{I}_p \otimes \boldsymbol\Sigma_k)$. We assign the skewness coefficients $\boldsymbol\psi_k$ a conjugate $\text{N}_J(\mathbf{m}_k,\mathbf{P}_k)$ prior. However, the updates of $\boldsymbol\beta_k$ and $\boldsymbol\psi_k$ can be combined into one step by defining the $(Jp + J) \times 1$ vector $\boldsymbol\beta^*_k = (\boldsymbol\beta_k,\boldsymbol\psi_k)^T$ for which we assume a $\text{N}_{Jp+J}(\mathbf{b}^*_k,\mathbf{I}_{(p+1)} \otimes \boldsymbol\Sigma_k)$ prior. For the multinomial logit model component, the regression parameters $\boldsymbol\delta_k$ are assigned a $\text{N}_r(\mathbf{d}_k, \mathbf{S}_k)$ prior for $k = 1,...,K-1$, which is conditionally conjugate as described in Section (\textbf{what section}).

We allow the normal-inverse-Wishart and multinomial hyperparameters to vary across clusters, though they may be shared across clusters in practice. An advantage of allowing for cluster-specific prior parameters is that \textit{a priori} knowledge of motor development trends can be incorporated into certain clusters while allowing the priors for other clusters to be less informative. Additionally, prior information regarding the effect of certain covariates on cluster membership can be incorporated in to the model by choosing informative values for $\mathbf{d}_k \ \text{and} \ \mathbf{S}_k$

\subsubsection{Posterior Inference} 
The above prior specification induces closed-form full conditionals that can be efficiently updated as part of a Gibbs sampler outlined below. A programatic sketch of our MCMC algorithm is given in Table (\ref{alg:MCMC}). Additional details, including derivations can be found in the Web Appendix.

@@

\paragraph{\textbf{Step 2:} Update of MSN Regression Parameters} We begin the update of MSN regression parameters by updating $t_i$, the truncated Normal random effect used in the stochastic representation of $\mathbf{y}_i$ given in equation \ref{eq:msnreg}. For cluster $k$, compute $A_k = (1 + \boldsymbol\psi_k^T \boldsymbol\Sigma^{-1}_k\boldsymbol
\psi_k)^{-1}$ according to current values of $\boldsymbol\psi_k$ and $\boldsymbol\Sigma^{-1}_k$. Now, for $i = 1,...,n_k$, where $n_k$ is the number of members of cluster $k$ according to the current values of ${z}_1,...,{z}_n$, compute $a_i = A_k \boldsymbol\psi_k^T \boldsymbol\Sigma^{-1}_k(\mathbf{y}_i - \boldsymbol\zeta_{ik})$. Finally, update draw $t_i$ from $\text{N}_{[0,\infty)}(a_i, \mathbf{A}_k)$. \textbf{Derivations in FS-P 2010 supplement}.
